---
title: Exploratory Data Analysis(EDA) for Alzheimer disease data
author: Chaeeun Shin
output: pdf_document
---
# Introduction
The purpose of this document is to get insight of the data before assumption, focusing on the patient's demographic details. 

# 1. Package Install
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```
1. *tidyverse*: This package is used for data manipulation.
2. *dplyr*: This package is used for data manipulation.
3. *ggplot2*: This package is used for data visualization.

# 2. Reading data
```{r}
data <- read.csv("alzheimers_disease_data.csv")
head(data)
```
# 3. Conversion to categorical data
```{r}
data$Gender <- as.factor(data$Gender)
data$Ethnicity <- as.factor(data$Ethnicity)
data$EducationLevel <- as.factor(data$EducationLevel)
data$Smoking <- as.factor(data$Smoking)
data$FamilyHistoryAlzheimers <- as.factor(data$FamilyHistoryAlzheimers)
data$CardiovascularDisease <- as.factor(data$CardiovascularDisease)
data$Diabetes <- as.factor(data$Diabetes)
data$Depression <- as.factor(data$Depression)
data$HeadInjury <- as.factor(data$HeadInjury)
data$Hypertension <- as.factor(data$Hypertension)
data$MemoryComplaints <- as.factor(data$MemoryComplaints)
data$BehavioralProblems <- as.factor(data$BehavioralProblems)
data$Confusion <- as.factor(data$Confusion)
data$Disorientation <- as.factor(data$Disorientation)
data$PersonalityChanges <- as.factor(data$PersonalityChanges)
data$DifficultyCompletingTasks <- as.factor(data$DifficultyCompletingTasks)
data$Forgetfulness <- as.factor(data$Forgetfulness)
data$Diagnosis <- as.factor(data$Diagnosis)

data <- data %>% dplyr::select(-c(PatientID, DoctorInCharge))
```
There are some categorical variables mis-classified as numeric data. They are assigned as factor. Also, the identification numbers of patients and doctors are removed since they are not used for any investigations for this project.

# 4. Missing data
```{r}
missing_summary <- sapply(data, function(x) sum(is.na(x)))
cat("\nMissing Data Summary\n")
print(missing_summary)
saveRDS(data,'data.rds')
```
There is no missing values in the dataset. 

# 5-1. Distribution of numeric variables
```{r}
cat("\nHistograms of Continuous Variables:\n")
continuous_columns <- data[, sapply(data, is.numeric)]
width <- c(Age = 10,BMI = 5,AlcoholConsumption = 5,PhysicalActivity = 2,DietQuality = 2,SleepQuality = 2,SystolicBP = 10,DiastolicBP = 10,CholesterolTotal = 30,CholesterolLDL = 30,CholesterolHDL = 10,CholesterolTriglycerides = 70,MMSE = 5, FunctionalAssessment = 2,ADL = 2)
for (col in colnames(continuous_columns)) {
  binwidth <- width[[col]]
 p <- ggplot(data, aes_string(x=col)) +
   geom_histogram(binwidth = binwidth, fill = 'skyblue', color = 'white') +
   labs(title = paste('Distribution of', col),
        x = col,
        y = "Frequency") +
   theme_minimal()
 print(p)
}
```
Histograms of numeric data are generated. 

# 5-2. Distribution of numeric variables, based on diagnosis status
```{r}
data_vis <- data %>%
  mutate(
    Diagnosis = factor(
      Diagnosis,
      levels = c(0, 1),
      labels = c("Does Not Have Alzheimer's", "Has Alzheimer's")
    )
  )

cat("\nDensity Plots of Continuous Variables by Diagnosis:\n")

for (col in colnames(continuous_columns)) {
  p <- ggplot(data_vis, aes(x = .data[[col]], fill = Diagnosis)) +
    geom_density(alpha = 0.5) +
    labs(title = paste("Density of", col, "by Diagnosis"), x = col) +
    scale_fill_manual(values = c("skyblue", "deepskyblue4")) +
    theme_minimal()
  print(p)
}
```
For better visualization, distribution of numeric variables are re-generated based on diagnosis status of Alzheimer's disease.

# 6-1. Distribution of categorical variables
```{r}
cat("\nHistograms of Categorical Variables:\n")
categorical_variables <- data[sapply(data, is.factor)]
categorical_variables$Gender <- ifelse(categorical_variables$Gender==0,'Male','Female')
categorical_variables <- categorical_variables %>%
  mutate(Ethnicity = case_when(
    Ethnicity == 0 ~ 'Caucasian',
    Ethnicity == 1 ~ 'African American',
    Ethnicity == 2 ~ 'Asian',
    TRUE ~ 'Other'),
    EducationLevel = case_when(
      EducationLevel == 0 ~ 'None',
      EducationLevel == 1 ~ 'High School',
      EducationLevel == 2 ~ "Bachelor's",
      TRUE ~ 'Higher')
  )

categorical_variables[,4:18] <- ifelse(categorical_variables[4:18]==0,'No','Yes')

for (c in 1:ncol(categorical_variables)) {
  counts <- categorical_variables %>% 
    count(!!sym(colnames(categorical_variables)[c])) %>%
    mutate(Percentage = n / sum(n) * 100)
  p <- ggplot(counts, aes_string(x=colnames(categorical_variables)[c], y = "n")) +
    geom_bar(stat = "identity", fill = "skyblue", color = "black") +
    geom_text(aes(label = paste0(round(Percentage,1),"%")), vjust = -0.5) +
    labs(title = paste("Bar Plot of", colnames(categorical_variables)[c]),
         x = colnames(categorical_variables)[c],
         y = "Count") + theme_minimal()
  print(p)
}
```
Frequency tables of categorical data are generated. 

# 6-2. Distribution of ctegorical variables, based on diagnosis status
```{r}
cat("\nHistograms of Categorical Variables based on Diagnosis:\n")
for (c in 1:ncol(categorical_variables)) {
  if (colnames(categorical_variables)[c]=='Diagnosis') next
  counts <- categorical_variables %>%
    count(!!sym(colnames(categorical_variables)[c]), Diagnosis)
  p <- ggplot(counts, aes_string(x = colnames(categorical_variables)[c], y = "n", fill = "Diagnosis")) +
    geom_bar(stat = "identity", position = "stack", color = "black") +
    labs(title = paste("Stacked Bar Plot of", colnames(categorical_variables)[c], "by Diagnosis"),
         x = colnames(categorical_variables)[c],
         y = "Count",
         fill = "Diagnosis") +
    theme_minimal() +
    scale_fill_manual(values = c("Yes" = "skyblue", "No" = "deepskyblue4"))  
  print(p)
}
```
For better understanding, frequency tables of categorical data are re-generated based on diagnosis status of Alzheimer's disease. 